---
title: "Extract 2D Road Geometry from OpenStreetMap (OSM)"
author: "Xiaoming Yang"
date: "2023-12-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This article introduces the procedure of using R to extract 2D road geometry data from OpenStreetMap and then plot the result using the 'leaflet' package. To extract roadway geometry data, we to create a overpass query using the 'osmdata' package.

First, load the required packages.

```{r message=FALSE}
library(osmdata)
library(leaflet)
```

The example code below extracts the geometry of I-75 in Bartow County, Georgia. Since I-75 is the only separated, controlled-access highway in the Bartow County, it is sufficient to extract 'motorway' features from the query. The overpass query returns a list 'opq' which includes linestring and point features of the roadway. Here we plot the linestring feature together with the bounding box of the county.  

```{r}
bbox <- getbb("Bartow County, Georgia") 
opq <- bbox %>% opq() %>%
  add_osm_feature(key = "highway", value = c("motorway")) %>%
  osmdata_sf()
roads <- opq$osm_lines

leaflet() %>%
  addTiles() %>%
    addPolylines(data = roads, color = "black", weight = 2) %>%
    addRectangles(
      lng1 = bbox[1], lat1 = bbox[2], lng2 = bbox[3], lat2 = bbox[4],
      weight = 2, color = "purple", fillColor = "transparent") %>%
    addScaleBar(position = "bottomleft")


```

Note that the result includes extra road segments beyond the bounding box. If the exact cut at the county border is desired, use the 'st_intersection()' function in the 'sf' package.

```{r message=FALSE, warning=FALSE}
library(sf)

bbox <- getbb("Bartow County, Georgia")
county <- getbb("Bartow County, Georgia",format_out = "sf_polygon")

opq <- bbox %>% opq() %>%
  add_osm_feature(key = "highway", value = c("motorway")) %>%
  osmdata_sf()
roads <- opq$osm_lines %>%  st_intersection(county)

leaflet() %>%
  addTiles() %>% 
  addPolylines(data = roads, color = 'black', weight = 2) %>%
  addPolygons(data = county,
                weight = 2,color = 'purple', fillColor = 'transparent') %>%
  addScaleBar(position='bottomleft')


```