---
title: "Horizontal Analysis"
author: "Xiaoming Yang"
date: "2023-12-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This article demonstrate the 'r-road-geometry project' in horizontal alignment analysis. The sample road used in this demonstration is the I-75NB in Bartow County. To run the code on this papge the function files 'fun_get_routes().R' and 'fun_horizontal_analysis.R' should be copied to the current working directory of R.

First, load required packages and function files.
```{r message=FALSE, warning=FALSE}
library(osmdata)
library(leaflet)
library(sf)

source('fun_get_routes.R') #identify unique routes
source('fun_horizontal_analysis.R') #horizontal alignment analysis
```


The following code extracts the 2D road geometry from OpenStreetMap(OSM).

```{r message=FALSE, warning=FALSE}
bbox <- getbb("Bartow County, Georgia")
county <- getbb("Bartow County, Georgia",format_out = "sf_polygon")

opq <- bbox %>% opq() %>%
  add_osm_feature(key = "highway", value = c("motorway")) %>%
  osmdata_sf()
roads <- opq$osm_lines %>%  st_intersection(county)

routes <- get_routes(roads)
route <- routes[2,]

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%  
  addPolylines(data = route, color = palette(rainbow(nrow(routes))), weight = 3) %>%
  addScaleBar(position='bottomleft')

```

Call the 'horizontal_analysis()' function. This function returns a sfc with all horizontal curves and tangent segments. These segments can be seperated into two collections: horizontal curves (hc) and horizontal tangents (ht). 

```{r}
hr<-horizontal_analysis(route)
hc <- hr %>% filter(nmpd < 0.05) #h curves
ht <- hr %>% filter(is.na(nmpd)) #h tangents

#Plot curves with best fit circles
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(data=hc, color='blue', weight = 4) %>%
  addPolylines(data=ht, color='black', weight = 2) %>%
  addCircles(data=hc,lng = ~lngc, lat = ~latc, radius = ~rad, color = "red",  
             weight = 1,fillColor = "transparent") %>%
  addScaleBar(position='bottomleft')

#Plot tangents
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(data=hc, color='blue', weight = 2) %>%
  addPolylines(data=ht, color='black', weight = 4) %>%
  addScaleBar(position='bottomleft')
```

A data summary can be easily obtained from R. 

```{r}
summary(hc)
summary(ht)
```


