---
title: "Vertical Alignment Analysis"
author: "Xiaoming Yang"
date: "2023-12-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This article demonstrate the 'vertical_analysis' function. The sample road used in this demonstration is the I-75NB in Bartow County. To run the code on this page, the function files 'fun_get_elevation().R' and 'fun_vertical_analysis.R' should be copied to the current working directory of R. Details about these functions can be found in previous articles. 


First, load required packages and function files.
```{r message=FALSE, warning=FALSE}
library(osmdata)
library(leaflet)
library(sf)
library(strucchange)
library(forecast)

source('fun_get_routes.R')
source('fun_get_elevation.R') #identify unique routes
source('fun_vertical_analysis.R') #horizontal alignment analysis
```


The following code extracts the 2D road geometry from OpenStreetMap(OSM).

```{r message=FALSE, warning=FALSE}
bbox <- getbb("Bartow County, Georgia")
county <- getbb("Bartow County, Georgia",format_out = "sf_polygon")

opq <- bbox %>% opq() %>%
  add_osm_feature(key = "highway", value = c("motorway")) %>%
  osmdata_sf()
roads <- opq$osm_lines %>%  st_intersection(county)

routes <- get_routes(roads)
route <- routes[2,]

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%  
  addPolylines(data = route, color = palette(rainbow(nrow(routes))), weight = 3) %>%
  addScaleBar(position='bottomleft')

```

For demonstration, we will focus on the beginning portion (~11.6km) of the route. The elevation data is extracted using the 'get_elevation()' function at 20m resolution.

```{r fig.height=5, fig.width=20, message=FALSE, warning=FALSE}
route2 <- route %>% 
          st_coordinates() %>% 
          head(150) %>% st_zm () %>% 
          st_linestring() %>% 
          st_sfc() %>% 
          st_set_crs(4326)


routes_xyz <- get_elevation(route2, 20)

vr <- vertical_analysis(routes_xyz) 

```

The output of the vertical_analysis() function is only the segmentation result with break points.We should apply curve and tangent criteria to identify vertical curves and tangents.

```{r fig.height=5, fig.width=20, message=FALSE, warning=FALSE}
vr <- vr %>% mutate(type = ifelse(r2 >= 0.5, "vc",
  ifelse(r2 <= 0.3 & std <= 1, "vt", "u")))
```

Finally, plot the result.

```{r fig.height=5, fig.width=20, message=FALSE, warning=FALSE}
coords <- routes_xyz %>% st_coordinates()
len_tot <- st_length(routes_xyz) %>% as.numeric()
df <- data.frame(
  dist = (c(1:nrow(coords)) - 1) / (nrow(coords) - 1) * len_tot,
  elev = coords[, 3]
)

plot(df,
  xlab = "Distance (m)", ylab = "Elevation (m)", type = "o", col = "blue",
  cex.lab = 1.5, cex.axis = 1.5,
  ylim = c(200, 300)
)
abline(v = df[vr[-1, ]$from, ]$dist, lty = "dashed")

for (i in 1:nrow(vr)) {
  if (vr[i, ]$type == "vc") {
    lines(x = c(df[vr[i, ]$from, ]$dist, df[vr[i, ]$to, ]$dist), y = c(207, 207), col = "red", lwd = 4)
  }
  if (vr[i, ]$type == "vt") {
    lines(x = c(df[vr[i, ]$from, ]$dist, df[vr[i, ]$to, ]$dist), y = c(205, 205), col = "black", lwd = 4)
    text(x = df[vr[i, ]$from, ]$dist + 100, y = 222, paste("G = ", round(vr[i, ]$avg, 2), "%"), srt = 90)
  }
}
legend(10500, 300, lty = 1, lwd = 4, col = c("red", "black"), 
       legend = c("Vertical curve", "Vertical tangent"), cex = 1)
```
Show vertical curves and tangents on a map.

```{r}
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(data=vr[vr$type=="vc",]%>%st_zm(), color='red', weight = 4) %>%
  addPolylines(data=vr[vr$type=="vt",]%>%st_zm(), color='black', weight = 2) %>%
  addScaleBar(position='bottomleft')

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(data=vr[vr$type=="vc",]%>%st_zm(), color='red', weight = 2) %>%
  addPolylines(data=vr[vr$type=="vt",]%>%st_zm(), color='black', weight = 4) %>%
  addScaleBar(position='bottomleft')
```
```


